Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-28 sources/bmc_template_submission/bmc_article.tex
final edits to figure caption

Rigorous analyses of multiple loci allow for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.
For example \cite{zhu_complexity_2021} show that when applying coalescent models,
there is more information in two genes of 300 bp each than in 600 independent sites.
Physcraper is desighed to facilitate
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral} or SVD Quartets \cite{chifman2014quartet}.
We have included a script (``multi\_locus.py'') and examples of how to automatically
merge the outputs of Physcraper runs at different loci into input files for these
two software, or as concatenated alignments for supermatrix analyses.

@@

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
Taxonomic links, and comparisons to existing published phylogenies in the OpenTree
data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses,
and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences.
The estimated gene tree itself is an evolutionarily explicit way to visualize gene
evolution, which in concert with taxonomic labelling can reveal paralogy.
Our conflict analysis tool informs the users of whether their tree contains major
conflicts with established taxonomy.
This tool also returns information on taxonomic conflicts that exist
in the original input tree.
Detected conflicts may be a sign that taxonomy needs to be updated, or may be a sign
that non-homologous sequences have been included in the analysis.
These taxonomic and phylogenetic conflicts flag regions of the tree for the researcher
to more closely examine and assess homology.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper from original
  \cite{gottlieb2005molecular} tree.
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and golden, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences in the original alignment that generated
  tree in B), filtered them following criteria from section "DNA sequence search and
  filtering", aligned them to original alignment using MUSCLE and performed a
  phylogenetic reconstruction using RAxML with 100 bootstraps.}
\end{figure}

>>>>

In the era of phylogenomics, rigorous analyses of multiple loci still allow for
more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.
For example \cite{zhu_complexity_2021} show that when applying coalescent models,
there is more information in two genes of 300 bp each than in 600 independent sites.
Physcraper is desighed to facilitate
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral} or SVD Quartets \cite{chifman2014quartet}.
We have included a script (``multi\_locus.py'') and examples of how to automatically
merge the outputs of Physcraper runs at different loci into input files for these
two software, or as concatenated alignments for supermatrix analyses.

@@

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
Taxonomic links, and comparisons to existing published phylogenies in the OpenTree
data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses,
and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences.
The estimated gene tree itself is an evolutionarily explicit way to visualize gene
evolution, which in concert with taxonomic labelling can reveal paralogy.
OpenTree's conflict analysis tool informs the users of whether their tree contains major
conflicts with established taxonomy and any phylogenetic context they wish to compare to.
This tool also returns information on taxonomic and phylogenetic conflicts that exist
in the original input tree.
Detected conflicts may be a sign that taxonomy needs to be updated, or may be a sign
that non-homologous sequences have been included in the analysis.
These taxonomic and phylogenetic conflicts flag regions of the tree for the researcher
to more closely examine and assess homology.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper using a starting phylogeny and alignment from
  \cite{gottlieb2005molecular} (Gottlieb2005 data in text).
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences in the Gottlieb2005 internal transcribed
  spacer DNA region (ITS) alignment,
  filtered them following criteria described in section "DNA sequence search and
  filtering", aligned them to the original alignment using MUSCLE, and performed a
  phylogenetic reconstruction using RAxML, with 100 bootstraps. B) Results of conflict
  analysis performed using OpenTree's conflict tool \cite{redelings2017supertree}.
  The Physcraper updated Gottlieb2005 phylogeny in (A) was compared to an \textit{Ilex}
  OpenTree synthetic subtree v. 12.3 \cite{ott727571} constructed using taxonomy of the genus
  as backbone and resolving branches based on phylogenetic data from the original
  Gottlieb2005 phylogeny.
\end{figure}
